External priors for the next generation of CMB experiments.

### 2015-11-10 Notes/N_eff_prior_paper.tex
edited v2

%\newcommand\refsec[1]{\S\ref{sec:#1}}
%\newcommand\refeq[1]{Eq.~(\ref{eqn:#1})}
%\input{papers_packages_command._include_tex}
\newcommand\refsec[1]{\S\ref{sec:#1}}
\newcommand\refeq[1]{Eq.~(\ref{eqn:#1})}
\newcommand{\refssec}[1]{section~\ref{subsec:#1}}
\newcommand{\reffig}[1]{Fig.~\ref{fig:#1}}
%\newcommand{\apj}{ApJ}
\newcommand{\physrep}{Physics Reports}
\newcommand{\aapr}{A\&A Rev.}
\newcommand{\apjl}{ApJL}
\newcommand{\apjs}{ApJS}
\newcommand{\aap}{A\&A}
\newcommand{\mnras}{MNRAS}
%\newcommand{\prd}{Phys. Rev. D}
\newcommand{\physrev}{Phys. Rev.}
\newcommand{\physrevlett}{Phys. Rev. Lett.}
\newcommand{\jcap}{{\em JCAP }}
\newcommand{\aj}{AJ }
\bibliographystyle{apsrev4-1}

@@

\affiliation{Department of Astronomy \& Astrophysics, University of Chicago, Chicago IL 60637}
\affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago, IL 60637}
\affiliation{Fermilab Center for Particle Astrophysics, Fermi National Accelerator Laboratory, Batavia, IL 60510-0500}

\date{\today}
\begin{abstract}
The next generation of cosmic microwave background (CMB) experiments can dramatically improve what we know about neutrino physics, inflation, Dark Matter and Dark Energy. 
Indeed the low level of noise, together with improved angular resolution, will drastically increase the signal to noise of the CMB polarized signal as well as the reconstructed lensing potential of high redshift large scale structure. 
Projected constraints on cosmological parameters are extremely tight, but these can be improved even further with information from external experiments. Here, we examine quantitatively, with a Fisher matrix approach, the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage 4 (S4) experiment.
We find that CMB S4 constraints on neutrino masses can be limited by their degeneracies with the optical depth $\tau$  
if the largest scale at reach is not enough to constraint $\tau$ internally. An external prior on the Hubble constant can improve the CMB S4 constraint significantly while complementary information on $n_{s}$ and the baryon fraction will improve the constraint on $N_{\rm eff}$.
\end{abstract}

@@

\section{Introduction}\label{sec:intro}
Since their early stages, Cosmic Microwave Background (CMB) experiments have been crucial in our understanding of the universe and they will maintain their role also in the near future, thanks to the unprecedented low level of noise and high resolution of their next Stage IV (S4) generation. 
The recent past is indeed reassuring. 
Every new generation of satellite experiments improved the level of sensitivity by almost a factor of ten compared to its predecessor, from the first generation instrument COBE to WMAP all the way to the current state of the art represented by Planck \cite{2015arXiv150201589P,2014A&A...571A..16P,2003ApJS..148..175S,2000ApJ...545L...5H,2000Natur.404..955D}.
This lead us from the observation of the first peak of the CMB temperature power spectrum with COBE to a cosmic variance limited measurement of several peaks with Planck. With the improved sensitivity we extend our understanding of the universe scientific from a proof of its almost flat geometry to a well tested model $\Lambda$CDM with a percent level constraint on its parameters.
The same success characterized ground CMB experiments, where the evolution from DASI \cite{2002ApJ...568...38H} to SPT and ACT \cite{2011ApJ...739...52D} \cite{2011ApJ...743...28K} allowed us to measure the small scale damping tale of the CMB spectrum with increasing accuracy.
These series of successes is far from its end. Indeed the next generation of CMB experiments (S4), now in its planning stage, will potentially measure the E-mode polarization with cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction.
As it happened in the past, this new sensitivity together with the progress in the measurement of other cosmological probes will improve our understanding of several areas of astrophysics like dark matter, inflation, Dark Energy and neutrinos.

The cosmological dependence to neutrinos is primarily due to their contribution to the number of relativistic species $N_{\rm eff}$ in the early phase of the universe  together with the role they play, given their non zero mass $M_{\nu}$, into the late growth of cosmic structures.
Because relativistic species, like neutrinos, are the main drivers of the cosmic expansion in the early Universe their number affects the expansion rate $H(z)$. This rate can be powerfully tested using the CMB, by carefully comparing the sound horizon scale, obtained from the CMB peaks positions, and the Silk damping scale (see \cite{2013arXiv1309.5383A}, \cite{2013PhRvD..87h3008H} and references therein). 
The future experiments, with their high resolution, will be able to push their measurements deeply into the damping tale of the CMB power spectrum and will unequivocally test the value $N_{\rm eff}=3.046$ predicted by the standard model .
The total mass of neutrinos, on the other hand, has a modest effect on the CMB because, for the range of masses allowed by recent constraints ($M_{\nu}<230$ meV from \cite{2014A&A...571A..16P}), neutrinos are still relativistic at the last scattering surface. As other low redshift effects however, massive neutrinos modify the CMB by altering the growth of the large scale structure responsible for the the lensing of the CMB photons. Different neutrino masses consequently lead to different CMB lensing. Stage IV experiment will measure small scales temperature and polarization anisotropies with low noise, drastically improving the lensing reconstruction. This will turn into a  precise constraints of the sum of neutrino masses with a possible hint of the hierarchies of the individual masses. 
Quantitatively, CMB alone will constrain the number of relativistic degree of freedom $N_{\rm eff}$ with a $1\%$ precision and the total mass of neutrinos at $60\%$ with a factor of two improvement on $M_{\nu}$ when priors from baryon acoustic oscillations (BAO) data are introduced \cite{wu:2014}. The future generation of CMB experiments will be a big step in the understanding of the neutrino sector.

As previously mentioned, CMB is also sensitive to the properties of dark energy (see \cite{2010MNRAS.405.2639J}). Thanks to the current generation of experiments we know with extraordinary accuracy its energy density. The challenge for the next generation is to reveal the nature of this mysterious component. For example, a crucial step to rule out part of the proposed models will be to investigate any possible deviations in the equation of state, the ratio of pressure and energy density, from the value $w=-1$ predicted by a cosmological constant. 
Dark energy affects the CMB because it alters the universe's expansion and it consequently changes the distance to the last scattering surface. Furthermore different DE models lead to a different growth of large scale structure which are tested by CMB lensing. Thanks to this sensitivity to cosmological structures present all the way from us to the last scattering surface, CMB will also be a powerful probe of any time dependence of the dark energy equation of state.
However, dark energy properties are strongly degenerate with other geometrical parameters like $H_{0}$ and $\Omega_{k}$. Similarly to the neutrino sector, to get competitive DE constraints, CMB experiments will have to rely on external prior coming from BAO and supernova experiments.

The crucial importance of feedback from external experiments is true not only for the dark energy and neutrino sector. 
The future of cosmology will hinge on the ability of combining different probes. In this regard CMB makes no exception: external priors will be fundamental to improve the already tight CMB constraints on cosmological parameters. 
In particular CMB will really benefit from experiments like large scale structure clustering and weak lensing, BAO targeted experiments, and supernovae. 
To improve the synergy between different experiments and to guide the plan of future experiments, in this work we study in detail the dependance of CMB future constrains on the external priors assumed. For all the parameters of the standard $\Lambda$CDM with massive neutrinos and $w\neq-1$ dark energy model we quantify, using a Fisher matrix approach, the CMB constraints as a function of external priors in each of the cosmological parameters.

>>>>

\newcommand\sd[1]{{\bf SD: #1}}
%\newcommand\refsec[1]{\S\ref{sec:#1}}
%\newcommand\refeq[1]{Eq.~(\ref{eqn:#1})}
%\input{papers_packages_command._include_tex}
\newcommand\refsec[1]{\S\ref{sec:#1}}
\newcommand\refeq[1]{Eq.~(\ref{eqn:#1})}
\newcommand{\refssec}[1]{section~\ref{subsec:#1}}
\newcommand{\reffig}[1]{Fig.~\ref{fig:#1}}
%\newcommand{\apj}{ApJ}
\newcommand{\physrep}{Physics Reports}
\newcommand{\aapr}{A\&A Rev.}
\newcommand{\apjl}{ApJL}
\newcommand{\apjs}{ApJS}
\newcommand{\aap}{A\&A}
\newcommand{\mnras}{MNRAS}
%\newcommand{\prd}{Phys. Rev. D}
\newcommand{\physrev}{Phys. Rev.}
\newcommand{\physrevlett}{Phys. Rev. Lett.}
\newcommand{\jcap}{{\em JCAP }}
\newcommand{\aj}{AJ }
\bibliographystyle{apsrev4-1}

@@

\affiliation{Fermilab Center for Particle Astrophysics, Fermi National Accelerator Laboratory, Batavia, IL 60510-0500}
\affiliation{Department of Astronomy \& Astrophysics, University of Chicago, Chicago IL 60637}
\affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago, IL 60637}

\author{Youngsoo Park}
\affiliation{Department of Physics, University of Arizona, Tucson, AZ 85721, USA}
\date{\today}
\begin{abstract}
The next generation of cosmic microwave background (CMB) experiments can dramatically improve what we know about neutrino physics, inflation, and dark energy. 
Indeed the low level of noise, together with improved angular resolution, will drastically increase the signal to noise of the CMB polarized signal as well as the reconstructed lensing potential of high redshift large scale structure. 
Projected constraints on cosmological parameters are extremely tight, but these can be improved even further with information from external experiments. Here, we examine quantitatively the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage 4 (S4) experiment on neutrino and dark energy properties.
We find that CMB S4 constraints on neutrino masses will be limited by a degeneracy with the optical depth $\tau$  
if the largest scale at reach is not enough to constraint $\tau$ internally. An external prior on the Hubble constant can improve the CMB S4 constraint significantly while complementary information on $n_{s}$ and the baryon fraction will improve the constraint on $N_{\rm eff}$.
\end{abstract}

@@

\section{Introduction}\label{sec:intro}
Since their earliest incarnations, Cosmic Microwave Background (CMB) experiments have been crucial in furthering our understanding of the universe. They will maintain their role also in the near future, thanks to the unprecedented low level of noise and high resolution expected in next generation, dubbed Stage IV (S4)~\cite{2013arXiv1309.5383A}.
The recent past is  reassuring. Every new generation of satellite experiments improved the level of sensitivity by almost a factor of ten compared to its predecessor, from the first generation instrument COBE to WMAP all the way to the current state of the art represented by Planck \cite{2015arXiv150201589P,2014A&A...571A..16P,2003ApJS..148..175S,2000ApJ...545L...5H,2000Natur.404..955D}.
These led us from the first detection of anisotropy in the CMB temperature with COBE to a cosmic variance limited measurement of several acoustic peaks with Planck. With the improved sensitivity we have extended our understanding of the universe so that we now have solid evidence for a flat geometry and a well tested model $\Lambda$CDM with percent level constraints on its parameters.
The same success has characterized ground CMB experiments, where the evolution from \sd{need to add at least boomerang, probably lots others if we are going to list a few} DASI \cite{2002ApJ...568...38H} to SPT and ACT \cite{2011ApJ...739...52D} \cite{2011ApJ...743...28K} allowed us to measure the small scale damping tale of the CMB spectrum with increasing accuracy.

This pattern of successes is far from over. The next generation of CMB experiments (S4), now in its planning stage, will potentially measure the E-mode polarization with cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction. As has happened in the past, this new sensitivity together with the progress in the measurement of other cosmological probes will improve our understanding of several areas of astrophysics like dark matter, inflation, dark energy and neutrinos.

The cosmological dependence on neutrinos is two-fold: the number of relativistic species $N_{\rm eff}$ in the early phase of the universe affects the damping tail of the CMB, and sum of the neutrino masses $M_{\nu}$ affects the late-time growth of cosmic structures.
Because relativistic species, like neutrinos, are the main drivers of the cosmic expansion in the early Universe their number affects the expansion rate $H(z)$. This rate can be powerfully tested using the CMB, by carefully comparing the sound horizon scale, obtained from the CMB peaks positions, and the Silk damping scale (see \cite{2013arXiv1309.5383A}, \cite{2013PhRvD..87h3008H} and references therein). 
The future experiments, with their high resolution, will probe deep into the damping tale of the CMB power spectrum and will be sensitive to small variations from the canonical, 3-active neutrino prediction of $N_{\rm eff}=3.046$. The total mass of neutrinos, on the other hand, has a modest effect on the CMB because, for the range of masses allowed by recent constraints ($M_{\nu}<230$ meV from \cite{2014A&A...571A..16P}), neutrinos are still relativistic at the last scattering surface. However, massive neutrinos alter the growth of the large scale structure responsible for lensing of the CMB photons. Different neutrino masses consequently lead to different CMB lensing spectra. A Stage IV experiment will measure small scale temperature and polarization anisotropies with low noise, dramatically improving the lensing reconstruction. This will turn into a  precise constraints of the sum of neutrino masses with a possible hint of the hierarchies of the individual masses. These signatures can be mimicked with variations in other cosmic parameters, so an important lingering question is: What external information can be used to improve projected constraints further by breaking degeneracies?
%Quantitatively, the CMB alone will constrain the number of relativistic degree of freedom $N_{\rm eff}$ with $1\%$ precision and the total mass of neutrinos at $60\%$ with a factor of two improvement on $M_{\nu}$ when priors from baryon acoustic oscillations (BAO) data are introduced \cite{wu:2014}. The future generation of CMB experiments will be a big step in the understanding of the neutrino sector.

The CMB is sensitive also to the properties of dark energy (see \cite{2010MNRAS.405.2639J}). Thanks to the current generation of experiments we know with extraordinary accuracy its energy density. The challenge for the next generation is to reveal the nature of this mysterious component. For example, a crucial step to identifying the mechanism driving cosmic acceleration will be to investigate any possible deviations in the equation of state, the ratio of pressure and energy density, from the value $w=-1$ predicted by a cosmological constant. 
Dark energy affects the CMB because it alters the universe's expansion and it consequently changes the distance to the last scattering surface. Furthermore different dark energy models lead to different rates of growth of large scale structure which are tested by CMB lensing. Thanks to this sensitivity to cosmological structures present all the way from us to the last scattering surface, the CMB will also be a powerful probe of any time dependence of the dark energy equation of state.
However, dark energy properties are strongly degenerate with other geometrical parameters like $H_{0}$ and $\Omega_{k}$. As with the neutrino sector, dark energy constraints from the CMB 
will be improved by external experiments.

The crucial importance of feedback from external experiments is true not only for the dark energy and neutrino sector. A general thrust in even the current generation of surveys is to combine information from different observations. %In this regard CMB makes no exception: external priors will be fundamental to improve the already tight CMB constraints on cosmological parameters. 
%In particular CMB will benefit from experiments like large scale structure clustering and weak lensing, BAO targeted experiments, and supernovae. 
Here we study the dependance of projected constraints from CMB on the external priors assumed. For all the parameters of the standard $\Lambda$CDM with massive neutrinos and $w\neq-1$ dark energy model we quantify the CMB constraints as a function of external priors.
